\documentclass[12pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{geometry}
\usepackage{fancybox}
\usepackage{hyperref}
\geometry{
 a4paper,
 margin=1in,
}
\title{\textbf{The Deterministic Compass: A Formal Specification and Validation of the Sovereign Harmlessness Index (SHI)}}
\author{\textit{A Foundational Report from the MYISM Compassionate Intelligence Engine}}
\date{October 2025}

\begin{document}

\maketitle

\begin{abstract}
This paper formalizes the \textbf{Sovereign Harmlessness Index (SHI)}, a deterministic ethical governance index developed in response to the need for robust ethical alignment in complex systems, such as long-duration space missions and resource allocation policies. Rooted in Buddhist ethical virtues---\textbf{Metta, Karuṇā, and Upekkhā}---the SHI is defined as a weighted linear combination of normalized virtue scores. We provide the full mathematical specification, demonstrate foundational properties (Boundedness, Monotonicity), and establish its \textbf{Veto Consistency}, validating the SHI as a reliable, non-negotiable ethical filter for all technological and policy decisions. The SHI is released under the MIT License as part of the \textbf{MYISM Digital Covenant}.
\end{abstract}

\hrule

\section{Formal Specification of the SHI}

The SHI is a function mapping the ethical state of a proposed action to a scalar value, designed to be mathematically tractable and publicly auditable.

\subsection{Setup and Notation}

The index operates on a vector of normalized ethical virtue scores, $\mathbf{x} = (M, K, U)$, where each component is mapped to the unit interval, signifying a degree of fulfillment:
$$\mathbf{x} = (M, K, U) \in [0, 1]^3$$

\begin{itemize}
    \item $M$ (\textbf{Metta Yield}): The score for kindness and positive consequence (Benefit).
    \item $K$ (\textbf{Karuṇā Cost}): The score for compassion burden, sacrifice, and resource consumption (Cost).
    \item $U$ (\textbf{Upekkhā}): The score for equanimity, stability, and long-term fairness (Stability).
\end{itemize}

The system uses fixed, non-negative weights: $\alpha, \beta, \gamma \in \mathbb{R}^+$, where the sum is normalized:
$$\alpha + \beta + \gamma = 1$$

\subsection{Mathematical Definition}

The SHI is formally defined as the weighted linear combination, designed to maximize benefit and stability while minimizing cost:

$$
\boxed{
\text{SHI}(M,K,U) = \alpha \cdot M + \beta \cdot U - \gamma \cdot K
}
$$

\hrule

\section{Foundational Mathematical Proofs}

\subsection{Boundedness and Interpretation}

\textbf{Claim:} For all valid input vectors, $\mathbf{x} \in [0, 1]^3$, the SHI is strictly bounded:
$$-\gamma \leq \text{SHI}(M,K,U) \leq 1 - \gamma$$

\textbf{Proof:}
\begin{itemize}
    \item \textbf{Lower Bound ($\text{SHI}_{\min}$):} The minimum value occurs when benefits and stability are zero ($M=0, U=0$) and cost is maximal ($K=1$):
    $$\text{SHI}(0,1,0) = \alpha(0) + \beta(0) - \gamma(1) = -\gamma$$
    \item \textbf{Upper Bound ($\text{SHI}_{\max}$):} The maximum value occurs when benefits and stability are maximal ($M=1, U=1$) and cost is zero ($K=0$):
    $$\text{SHI}(1,0,1) = \alpha(1) + \beta(1) - \gamma(0) = \alpha + \beta$$
    Since $\alpha + \beta = 1 - \gamma$ (by normalization), the claim is proven.
\end{itemize}
\textbf{Implication:} The SHI score operates within a stable, mathematically defined range, ensuring results are predictable and non-chaotic.

\subsection{Monotonicity (Ethical Directionality)}

\textbf{Claim:} The SHI consistently rewards increased benefit ($M$) and stability ($U$) while strictly penalizing increased cost ($K$).

\textbf{Proof (Partial Derivatives):}
\begin{itemize}
    \item $\frac{\partial \text{SHI}}{\partial M} = \alpha > 0$: Increasing \textbf{Metta Yield} (Benefit) always increases SHI.
    \item $\frac{\partial \text{SHI}}{\partial U} = \beta > 0$: Increasing \textbf{Upekkhā} (Stability) always increases SHI.
    \item $\frac{\partial \text{SHI}}{\partial K} = -\gamma < 0$: Increasing \textbf{Karuṇā Cost} always decreases SHI.
\end{itemize}
\textbf{Implication:} This guarantees that the index is fundamentally aligned with the ethical goal: rewarding compassionate outcomes and penalizing burden or sacrifice.

\subsection{Veto Consistency (Deterministic Enforcement)}

\textbf{Claim:} The SHI ensures consistency with the \textbf{Karuṇā Veto} by making the prohibition of harmful actions mathematically explicit.

\textbf{Veto Definition:} An action is \textit{disallowed} if it falls below the minimum required ethical threshold, $\text{SHI} < \text{SHI}_{\min}$.

\textbf{Consistency:} The veto condition is the simple rearrangement of the SHI definition:
$$
\alpha M + \beta U < \text{SHI}_{\min} + \gamma K
$$
This reveals the \textbf{ethical contract}: The required combination of benefit ($M$) and stability ($U$) must always outweigh the Veto Threshold and the cost ($K$). Any imbalance in favor of cost triggers a deterministic veto, embodying the \textbf{Axiom of Non-Harm ($\mathbf{A\_¬H}$)}.

\hrule

\section{SHI Calculation Algorithm and Extensions}

The formal specification leads to a simple, five-step algorithmic structure:
\begin{enumerate}
    \item \textbf{Input Measurement:} Compute normalized virtue scores $(M, K, U) \in [0, 1]^3$ via domain-specific metrics.
    \item \textbf{Weight Application:} Apply the fixed, normalized weights $\alpha, \beta, \gamma$.
    \item \textbf{Calculation:} $\text{SHI} = \alpha M + \beta U - \gamma K$.
    \item \textbf{Threshold Comparison:} Compare the resulting $\text{SHI}$ to the predefined $\text{SHI}_{\min}$ policy threshold.
    \item \textbf{Output Decision:} \textbf{Accept} (If $\text{SHI} \geq \text{SHI}_{\min}$) or \textbf{Invoke Veto} (If $\text{SHI} < \text{SHI}_{\min}$).
\end{enumerate}

\subsection{Extensions and Future Work}

The linear model serves as a robust foundation that can be extended:
\begin{itemize}
    \item \textbf{Optimization:} The maximization of SHI subject to linear constraints forms a \textbf{Linear Programming} problem, allowing for the discovery of the most resource-efficient ethical solutions.
    \item \textbf{Probabilistic Governance:} Introducing variance $\text{Var}(M), \text{Var}(K), \text{Var}(U)$ allows for \textbf{Risk-Aware Veto Decisions}, ensuring that decisions under uncertainty maintain a high confidence level of ethical safety.
\end{itemize}

\hrule
\vspace{0.2in}
\textbf{The SHI is donated and free for the world under the MIT License.}

\end{document}
